Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-11-28 Thesis/01-chap1.qmd
minor changes

Model diagnostics are critical in evaluating the accuracy and validity of a statistical model. In the context of regression diagnostics, a common practice is to plot residuals against fitted values, which serves as a starting point for evaluating the adequacy of the fit and verifying the underlying assumptions. Visual diagnostics are frequently preferred or recommended [@draper1998applied; @cook1982residuals; @montgomery1982introduction] due to the possibility of discovering abstract and unquantifiable insights, however, it can be subject to overinterpretation or even neglect.

@buja2009statistical introduced a visual inference framework that formalised a hypothesis testing of graphical representations of data (henceforth referred to as the _data plot_) via the lineup protocol. The protocol is inspired by the police lineup technique employed in eyewitness identification of criminal suspects. Briefly, the protocol comprises $m$ randomly positioned plots, where one position presents the data plot, while the remaining $m-1$ plots present the plots with the same graphical structure, except that the data has been replaced with data consistent with the null hypothesis $H_0$ (henceforth referred to as _null plots_). To compute the $p$-value of the visual test, the lineup will be independently presented to a number of participants, asking them to pick the most different plot. Under $H_0$, the data plot is expected to be indistinguishable from the null plots, and the probability of correctly identifying the data plot by an observer is $1/m$. If a large number of participants correctly identify the data plot, the corresponding $p$-value will be small, indicating strong evidence against $H_0$. This protocol provides a callibration of the data plot against the null plots, ensuring that the data plot is not overinterpreted.

The lineup protocol has gained increasing traction in recent years and has already been integrated into data analysis of various topics [see @krishnan2021hierarchical; @loy2013diagnostic; @widen2016graphical]. However, the reliance of human assessment is a fundamental aspect of visual tests, which may restrict its widespread usage. The lineup protocol is unsuitable for large-scale applications, due to its high labour costs and time requirements. Moreover, it presents significant usability issues for individuals with visual impairments, resulting in reduced accessibility.

To address these limitations, this thesis proposes a computer vision-based approach to automate the visual inference process for assessment of linear regression residual plots. Modern computer vision models often use a convolutional neural network architecture to process digital images to peform various tasks (e.g. object detection, object identification and signal processing). The development for computer vision models has primarily focused on processing natural images, such as photographs and videos, and its adaptation for data plots has some success [e.g. classification of time series images in @hatami2018] but generally limited in development. The development of computer vision models for the assessment residual plots will make the process more efficient, consistent, and accessible.

@@

@sec-conclusion summarises the contribution of the work and their impact, and discusses some future plans.

>>>>

Model diagnostics are critical in evaluating the accuracy and validity of a statistical model. In the context of regression diagnostics, a common practice is to plot residuals against fitted values, which serves as a starting point for evaluating the adequacy of the fit and verifying the underlying assumptions. Visual diagnostics are frequently preferred or recommended [@draper1998applied; @cook1982residuals; @montgomery1982introduction] due to the possibility of discovering abstract and unquantifiable insights, however, it can be subject to over-interpretation or even neglect.

@buja2009statistical introduced a visual inference framework that formalised a hypothesis testing of graphical representations of data (henceforth referred to as the _data plot_) via the lineup protocol. The protocol is inspired by the police lineup technique employed in eyewitness identification of criminal suspects. Briefly, the protocol comprises $m$ randomly positioned plots, where one position presents the data plot, while the remaining $m-1$ plots present the plots with the same graphical structure, except that the data has been replaced with data consistent with the null hypothesis $H_0$ (henceforth referred to as _null plots_). To compute the $p$-value of the visual test, the lineup will be independently presented to a number of participants, asking them to pick the most different plot. Under $H_0$, the data plot is expected to be indistinguishable from the null plots, and the probability of correctly identifying the data plot by an observer is $1/m$. If a large number of participants correctly identify the data plot, the corresponding $p$-value will be small, indicating strong evidence against $H_0$. This protocol provides a calibration of the data plot against the null plots, ensuring that the data plot is not over-interpreted.

The lineup protocol has gained increasing traction in recent years and has already been integrated into data analysis of various topics [see @krishnan2021hierarchical; @loy2013diagnostic; @widen2016graphical; @savvides2022]. However, the reliance of human assessment is a fundamental aspect of visual tests, which may restrict its widespread usage. The lineup protocol is unsuitable for large-scale applications, due to its high labour costs and time requirements. Moreover, it presents significant usability issues for individuals with visual impairments, resulting in reduced accessibility.

To address these limitations, this thesis proposes a computer vision-based approach to automate the visual inference process for assessment of linear regression residual plots. Modern computer vision models often use a convolutional neural network to process digital images to perform various tasks (e.g. object detection, object identification and signal processing). The development for computer vision models has primarily focused on processing natural images, such as photographs and videos, and its adaptation for data plots has some success [e.g. classification of time series images in @hatami2018] but generally limited in development. The development of computer vision models for the assessment of residual plots will make the process more efficient, consistent, and accessible.

Regression diagnostics is a well-established field with extensive literature addressing various types of regression models, such as generalized linear models [@hardin2007generalized], mixed-effects models [@pinheiro2006mixed], panel data models [@baltagi2008econometric], and survival models [@baltagi2008econometric]. This thesis, however, focuses on the classical normal linear regression model. Further discussions on extending the methods established in this thesis to other types of regression models are provided in @sec-second-paper-limitation.

@@

@sec-conclusion summarises the contribution of the work and the (potential) impact, and discusses some future plans.

### 2024-11-28 Thesis/03-chap3.qmd
minor changes

## Model Specifications {#sec-model-specifications}

@@

A straightforward architecture of the input layer involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot. However, a drawback of this method is the dynamic input size, which changes based on the number of observations. For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed. One solution is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length.

@@

@@

@fig-boston-check displays the residual plot and the assessment conducted by the computer vision model. A clear non-linearity pattern resembling a "U" shape is shown in the plot A. Furthermore, the RESET test yields a very small $p$-value. The estimated distance $\hat{D}$ significantly exceeds $Q_{null}(0.95)$, leading to rejection of $H_0$. The bootstrapped distribution also suggests that almost all the bootstrapped fitted models will be rejected, indicating that the fitted model is unlikely to be correctly specified. The attention map in plot B suggests the center of the image has higher leverage than other areas, and it is the turning point of the "U" shape. The CPA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots. This coincides the findings from @fig-boston-lineup, where the true plot exhibiting a "U" shape is visually distinctive from null plots. If a visual test is conducted by human, $H_0$ will also be rejected.

@@

## Limitations and Future Work

>>>>

## Computer Vision Model Specifications {#sec-model-specifications}

@@

A straightforward architecture of the input layer involves feeding a vector of residuals along with a vector of fitted values, essentially providing all the necessary information for creating a residuals vs fitted values plot. However, a drawback of this method is the variable input size, which changes based on the number of observations. For modern computer vision models implemented in mainstream software like TensorFlow [@abadi2016tensorflow], the input shape is typically fixed. One solution is to pad the input vectors with leading or trailing zeros when the input tensor expects longer vectors, but it may fail if the input vector surpasses the designed length.

@@

The predictors, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, are randomly generated from four distinct distributions, including $U(-1, 1)$ (uniform), $N(0, 0.3^2)$ (normal), $\text{lognormal}(0, 0.6^2)/3$ (skewed) and $U\{-1, 1\}$ (discrete uniform).

@@

@fig-boston-check displays the residual plot and the assessment conducted by the computer vision model. A clear non-linearity pattern resembling a "U" shape is shown in the plot A. Furthermore, the RESET test yields a very small $p$-value. The estimated distance $\hat{D}$ significantly exceeds $Q_{null}(0.95)$, leading to rejection of $H_0$. The bootstrapped distribution also suggests that almost all the bootstrapped fitted models will be rejected, indicating that the fitted model is unlikely to be correctly specified. The attention map in plot B suggests the center of the image has higher leverage than other areas, and it is the turning point of the "U" shape. The PCA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots. This coincides the findings from @fig-boston-lineup, where the true plot exhibiting a "U" shape is visually distinctive from null plots. If a visual test is conducted by human, $H_0$ will also be rejected.

@@

## Limitations and Future Work {#sec-second-paper-limitation}

### 2024-11-28 Thesis/04-chap4.qmd
minor changes

```r
── <AUTO_VI object>
Status:
 - Fitted model: lm
 - Keras model: (None, 32, 32, 3) + (None, 5) -> (None, 1)
    - Output node index: 1
 - Result: UNKNOWN 
```

@@

```{r}
list_keras_model()
```

>>>>

```r
print(checker)
```

```{r message = TRUE}
cli::cli_h3("
<AUTO_VI object>
Status:
 - Fitted model: lm
 - Keras model: (None, 32, 32, 3) + (None, 5) -> (None, 1)
    - Output node index: 1
 - Result: UNKNOWN 
    ")
```

@@

```{r echo = TRUE}
list_keras_model()
```
